Modelling BCG vaccination in the UK

### 2019-07-31 rmds/09-model-fitting.Rmd
model fitting captions updated
first spell check

In the previous chapter I outlined a mechanistic model of Tuberculosis (TB) transmission. Whilst this model made use of the best available evidence there remains a large degree of uncertainty regarding its structure and parameterisation. The majority of this uncertainty relates to the amount of TB transmission occuring in England. An approach to deal with this uncertainty is to fit the model to available observed data. Model fitting involves optimising over the available parameter space to return parameter sets that fit the data in some quantative way "better" than other parameter sets.

This chapter details an approach to challenging a infectious disease model to data using the state-space model formulation and bayesian model fitting techniques. It first outlines the infectious disease model discussed in the previous chapter as a state-space model, as well as detailing the data used for fitting the model, and the parameters that are fitted. It then outlines the theoretical, and practical, justification for the model fitting pipeline used to calibrate and fit this state space model. Finally it discusses the quality of the model fit, ad-hoc techniques used to improve model fit, strengths and limitations of the approach, and areas for further work.

@@

State space models (SSMs) may be used to model dynamic systems that are partially observed (such as all but the most contained infectious disease outbreak or endemic). They consist of a set of parameters, a latent continuous/discrete time state process and an observed continuous/discrete time process.[@Murray2015] The model developed in the previous section represents the state process of the SSM, with the parameters estimated for the model representing the model initial conditions and parameter set. To complete the mapping to an SSM an observational model is required. This observational model takes the latent estimates from the dynamic model and forecasts the observed data. We specifiy such an observational model in the Section \@ref(observational-model).

@@

The primary data source used was the reported UK born TB notifications from 2000 to 2004 as recorded by the Enhanced TB Survelliance (ETS) system (see Chapter \@ref(data)). These are the only years for which notifications are available stratified by UK birth status and for which universal school-age BCG vaccination was in place. As the model of TB transmission outlined in the previous chapter had multiple age-dependent parameters UK born TB notifications stratified by age group are also fitted to. The age groups considered are children (aged 0-14 years old), adults (15-69 years old), and older adults (70-89 years old). These age groups were used, rather than the more detailed 5 year age groups used in the TB transmission model (see Chapter \@ref(model-development)), for three reasons. Firstly, by using condensed age groups the number of notifications in each group increased. Having a larger number of notifications in each group reduces the impact of stochastics noise, which makes fitting the model easier. Secondly, reducing the number of data points, whilst still capturing the important age dynamics, reduces the compute requirements of the model (see Section \@ref(particle-filter)). As will be discussed in Section \@ref(fitting-pipeline) this was a major consideration as the model fitting approached used was highly compute intensive. Finally, the condensed age groups used were chosen to be meaningful. This allows model results to be more easily reported and interpreted.

@@

There are three major considerations to account for when developing an observed disease notification model (i.e a reporting model). These are: systematic reporting error over time; systematic changes in reporting error over time; and reporting noise. For the notification data outlined in the previous section there is little evidence to suggest the form that this measurement model should take. For this reason I have assumed that all reporting errors are gaussian and that their are no time variable reporting errors. The reporting model can be then defined as follows,

@@

Where $O$ are the observed notifications, $A$ are the incident cases of disease as forecast by the disease transmission model, $E_{\text{syst}}$ is the systematic reporting error, $E_{\text{noise}}$ is the reporting noise, and $\mathcal{N}$ represents the gaussian (normal) distribution. The priors for the model are defined in Table \@ref(tab:measurement-model). The prior for systematic reporting error is based on the assumption that cases are less likely to be reported than they are to be over reported. The prior for the reporting noise is an assumption. This observation model was also used for the non-UK born pulmonary cases.

A potential limitation of this model is that reporting of TB cases is likely to have improved over time. This is especially true of notifications reported prior to the introduction of the ETS in 2000 (see Chapter \@ref(data)). A potential improvement to this model would be to introduce seperate systematic reporting errors for notifications pre and post the introduction of the ETS. However, given that I have only included a single point prior to the introduction to the ETS doing so in this instance would lead to overfitting of these parameters. Instead I have assumed that $ E_{\text{noise}}$ is doubled for data points prior to the introduction of the ETS.

@@

The model outlined in the Chapter \@ref(model-development) has a large number of free parameters for which prior distributions have been specified based on the observed data, the literature, and expert knowledge. In theory the model fitting pipeline outlined below could be used to produce posterior distributions for all these parameters. However, in practise this is not feasible as the data discussed in Section \@ref(observed-data) only covers notifications and therefore does not contain sufficient information. If every parameter was allowed to update based on the data then it is likely that the resulting posterior distributions would not match with alternative data sources and the literature. Another potential issue is that by allowing all parameters to be fitted the meaningful information in the observed data may be lost.

@@

In addition for scenarios with age variable transmission probabilities or non-UK born mixing the following paramters may also be fitted to,

* Transmission probablity modifier for young adults ($\beta_{\text{young adult}}$).
* Non-UK born mixing modifier for young adults ($M_{\text{young adult}}$).

@@

Fitting dynamic transmisson models is complex and requires the use of specialist statistical techniques. There are a variety of these tools available. Ranging for tried and tested to cutting edge. Historically many modellers have used maximum likelihood methods to fit deterministic models. Unfortunately, whilst computationally efficient, these methods only provide point estimates and are limited to relatively simple models. More recently Bayesian methods have become popular. These have numerous benefits including: explicit inclusion of prior knowledge via prior distributions for all parameters; ability to handle complex stochastic models; and provide parameter distributions (posterior distribution) of best fitting parameters rather than single point estimates. Unfortunately these methods also require calibration prior to use. This section outlines the theoretical justification, and implementation details, of an automated model fitting pipeline used to calibrate, and fit, the previously detailed state space model of TB tranmission in England.

Libbi was used for all model fitting.[@Murray2015] LibBi is a software package for state-space modelling and Bayesian inference. It uses a domain specifc language for model specification, which is then optimised and compiled to provide highly efficient model code. It focusses on full information model fitting approaches including: particle Markov chain Monte Carlo (PMCMC), and SMC-SMC methods for parameter estimation. All fitting algorithms are highly efficient and scalable across mutliple CPUs or GPUs. `RBI` and `RBI.helpers` were used to interface with LibBi from R.[@Funk:2019ud; @Funk:2019uw] `RBI.helpers` was also used to optimise the model fitting pipeline as detailed in the calibration section. As model fitting using `LibBi` is compute intensive a workstation was built, and overclocked, with these compute requirements in mind^[See these blog posts for details: https://www.samabbott.co.uk/post/building-an-rstats-workstation/, https://www.samabbott.co.uk/post/benchmarking-workstation-xgboost/, https://www.samabbott.co.uk/post/benchmarking-workstation-benchmarkme/]. All model fitting code is available on GitHub as an R package^[See here for details: https://github.com/seabbs/ModelTBBCGEngland].

@@

In order to fit a model to data it is neccessary to estimate, or calculate, the marginal likelihood. Mathematically, the marginal likelihood is the plausibility that a parameter set, given the specified statistical model and the initial conditions, describes the observed data. For complex state space models, such as that discussed in the previous chapter, calculating the marginal likelihood is not possible.[@Murray2015] The particle filter provides a model-agnostic approach, based on importance sampling, to estimate the marginal likelihood. The variant used in this thesis, the bootstrap particle filter, is described below. See [@Murray2015] for a more technical discussion of the bootstrap particle filter.

@@

2. Sequentially for the each observed data point the particle filter is then advanced through a series of *propagation*, *weighting*, and *resampling* steps.

@@

The particle filter approach outlined above, is a member of a family of sequential monte carlo (SMC) methods. These methods all initialise particles and then follow the same *propagation*, *weighting*, and *resampling* steps as previously detailed. SMC may also be used to sample from the posterior distribution of a given set of priors and a specified model. This works as follows,

@@

2. These parameter particles are then iterated sequentially over each observed data point, undergoing the same *propagation*, *weighting*, and *resampling* steps as in the particle filter, as well as an additional *rejuvenation* step.

@@

   * *Weighting:* Parameter particles are weighted using the marginal likelihood. In principle this could be computed exactly, but is most commonly estimated using a nested particle filter for each state particle (i.e as outlined in the previous section). For a subset of models a Kalman filter maybe used instead.[@Murray2015] The marginal likelihood may also be estimated using other partial information techniques such as approximate bayesian computation. In the case where a particle filter is used the full algorithm is known as Sequential Monte Carlo - Sequential Monte Carlo (SMC-SMC). This algorithm is used for all dynamic model fitting in this thesis.

 *  *Resampling:* The parameter particle stock is restored to equal weights by resampling particles, with replacement, with the probability of each sample being drawn being proportional to its weight.

 * *Rejuvenation:* *Resampling* of the parameter particles at each time point leads to a reduction in the number of unique values present. For state particles (when estimating the marginal likelihood using a particle filter) particles are diversifed with each propagation but as parameters do not change in time parameter particles cannot diversify in this way. To account for this the *rejuvenation* step is inserted after the resampling of parameter particles at each time point. The *rejuvenation* step is a single, or multiple depending on the acceptance rate, Metropolis-Hastings step for each parameter particle. This step aims to preserve the distribution of the parameter particles, whilst increasing their diversity. To minimise unnessary rejuvenation an effective sample size thresold can be used. This only triggers rejuveation when particle diversity has decreased below the target effective sample size thresold.

@@

Where $y$ is the observed data, $\theta$ is the current parameter set, $\theta'$ is some proposed parameter set sampled from some proposal distribution $q(\theta' | \theta)$.[@Murray2015] By construction samples drawn using this rule are ergodic to the posterior distribution. This means that after convergance samples drawn using this rule may be considered as samples from the posterior distribution.

@@

The accuracy of the marginal likelihood estimate returned by the particle filter is dependent on the number of particles used, the number of observed data points, the parameter sample, and the complexity of the model. As the number of particles tends towards infinity the likelihood estimate provided by the particle filter should tend towards the exact solution. This suggests that choosing a very high number of particles maybe the optimal solution in terms of accuracy. Unfortunately, each particle requires a full model simulation, which for complex models can be computationally costly. This means that using very large numbers of particles is not tractable. For this reason it is neccassary to determine an optimal number of particles that both provides an adequately accurate estimate of the likelihood whilst being computationally tractable.

The `rbi.helpers` R package attempts to solve this issue by adopting the following strategy.[@Funk:2019uw] First, the approximate mean of the posterior distribution is obtained, as accurate likelihood estimates near the posterior mean are of the most interest. Repeated model simulations are then run using the same set of parameters, with the marginal likelihood being estimated each time using a given number of particles. The variance of these log-likelihood estimates is then calculated. This process is then repeated for increasing numbers of particles until the log-likelihood variance is below some target thresold, commonly 1.[@Funk:2019uw]

I have implemented this as a two step process for each fitted scenario. Firstly, I used the Nelder-Mead simplex method, via LibBi,[@Murray2015] to find a parameter set that optimised the maximum likelihood. I then initialised a 1000 step PMCMC chain with this parameter set, using 256 particles in the particle filter. I then used `rbi.helpers`,[@Funk:2019uw] as outlined above, to estimate the number of particles required to produce a log-likelihood variance of less than 1 for this sample of the posterior distribiution, using 250 samples per step and starting with 16 particles. I initially planned to repeat this process for multiple draws from the posterior distribution but this proved to be infeasible given the compute available. A target of 5 for the log-likelihood variance was chosen as a smaller target could not be feasibly achieved given the compute resources available. Additionally 256 was specified as the maximum number of feasible particles to use in the particle filter.

@@

When using an MCMC alogorithm a proposal distribution is required to provide new parameter samples to evaluate. For SMC-SMC a proposal distribution is required to inform the MCMC sampler that is run during the rejuvernation step. By default if no proposal distribution is provided Libbi uses the prior distribution.[@Murray2015] The prior distribution can be an inefficient proposal distribution as it is likely to have a low acceptance rate (from the MCMC sampler).[@Murray2015] Having a low acceptance rates means that many more MCMC steps are required to generate a successful parameter sample. This results in slow mixing and computationally expensive MCMC steps may make model fitting intractable.

A more efficient approach is to specify a proposal distribution that draws parameter samples that are closer to the current state of the MCMC chain than the overall prior distribution. There is an extensive literature examing how to optimise the proposal distribution to achieve an good acceptance rate. In practise it has been shown that a rate of between 10% and 20% is optimal for upwards of 5 parameters.[@Murray2015] This strikes a balance between allowing the chain to fully explore the posterior distribution whilst still being as efficient as possible.

A simple approach to setting the proposal is to run a series of MCMC steps and then calculate the acceptence rate. Based on the acceptence rate the width of the proposal distributions can then be adapted. By repeating these steps multiple times a proposal distribution which gives an acceptance rate within the desired bounds can be arrived at. This adaption can either be independent for each parameter or dependent (taking into account empirical correlations). The `adapt_proposal` function, from the `rbi.helpers` R package,[@Funk:2019uw] implements this approach and is used in this model fitting pipeline. In many models parameters are likely to have strong correlations (i.e between UK and Non-UK born mixing rate and effective contact rate) and in these scenarios it is likely that a dependent strategy for adapting the proposal distribution will more efficiently explore the posterior distribution. However, the downside of adapting the proposal distribution using dependent methods is that the resulting proposal is highly complex, is computationally expensive to compute and may breakdown in some areas of the posterior distribution.

In this model fitting pipeline I have used a maximum of 5 iterations of, manual, independent proposal adaption, drawing 250 samples in each iteration, starting with guassian distributions for each parameter, truncated by the range of the prior, with the mean based on the current parameter value. The standard deviation for each parameter was assumed to be the standard deviation of the prior if it was guassian and otherwise assumed to be the range of the prior if it was uniform. For each iteration I halved the  size of the standard deviation of each parameter. As for the particle calibration, I initially used a maximum likelihood method to provide a point estimate of the best fitting parameter set, followed by 1000 PMCMC steps, using a 256 particle filter. This means that the proposal distribution is adapted near to the posterior mean rather than in the tails of the posterior distribution.

I chose to use, manual, independent proposal adaption method for several reasons. Firstly, when developing this pipeline the approaches implemented in `rbi.helpers` produced multiple transient errors in other `rbi` and `rbi.helpers` code. Secondly, the resulting dependent proposal distribution was highly complex, slow to compute, and diffult to debug. Finally, for SMC-SMC efficient exploration of the proposal distribution is less important than when using MCMC alone as SMC-SMC is initialised with multiple samples from the prior distribution. This means that multiple local maximas can be efficiently explored regardless of the proposal distribution used. The MCMC rejuvernation step then serves to provide additional samples from these local maximas. Proposal adaption was only carried out for the main model scenario with all other scenarios using this proposal distribution.

### Model comparision

@@

Smaller DIC values should indicate a better fit to data than larger DIC values. The DIC is composed of the deviance, which favours a good fit, and the effective number of parameters, which penalises overfitting.[@Gelman:nll_LBlw] Unlike the AIC the DIC can be estimated using samples from the posterior distribution and so is more readily calculated for models estimated using bayesian methods. It can be defined as,

$$ {\mathit  {DIC}}=D({\bar  {\theta }})+2p_{D}$$
Where $\bar{\theta}$ is the expectation of $\theta$, with $\theta$ being defined as the unknown parameters of the model. $p_{D}$ is the effective number of parameters in the model and is used to penalise more comple complex models. It can be estimated as follows,[@Gelman:nll_LBlw]

@@

PRCC is a sampling based approach which can be computed with minimal computational cost from a sample of the prior or posterior distributions of a model. It estimates the degree of correlation between a given parameter input and an output after adjusting (using a linear model) for variation in other inputs. It is an extension of more simplistic sampling techniques, the most basic of which, is simply examining scatter plots of a sampled parameter set against the outcome of interest. PRCC is required as these more simplistic techniques become intractable with higher dimensionality as they do not account for between paremeter correlation or are just difficult to interpret with multiple dimensions.[@Marino2009a] PRCC can be understood by first outlining the individual steps. These are:

@@

Where $\operatorname {cov}$  is the covariance, $\sigma _{X}$ is the standard deviation of $X$, and $\sigma_Y$  is the standard deviation of $Y$. Where $X$ is the input and $Y$ is the output.

@@

A limitation of PRCC is that it whilst it can capture non-lnear relationships between outputs and inputs these relationships must be monotonic.[@Marino2009a] For relationships that are non-monotic methods that rely on the decompositon of model output variance, such as the extended Fourier amplitude sensitivity test,[@Marino2009a] are more appropriate. However, these approaches are computationally demanding as they typically require mutliple iterations of model simulation. Additionally, they cannot be used on a previous parameter samples, instead needing to sample and simulate the model within the parameter sensitivity algorithmn. This means that they cannot be used for "free" (i.e with neglible additional compute cost), unlike PRCC which can be estimated using a sample from the posterior distribution. For this reason these approaches have not been further explored in this thesis.

I have implemented PRCC using the `epiR` R package^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/test_sensitivity.R],[@EpiR] using the samples from the posterior distribution of the model calculated during the SMC-SMC step. Parameter sensitivity measures such as PRCC must be calculated seperately for each output time point. I calculated the PRCC for each fitted parameter, at the final time point with fitted data (2004), for overall TB incidence rates. These results are then summarised by plotting the absolute PRCC values, indicating the direction of correlation using colour^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/plot_sensitivity.R].

@@

1. Model initialisation using minimal error checking and single precision computation. Implemented using the `disable-assert` adn `enable-single` flags in LibBi.[@Murray2015] Outputs are only given for times with observed data and a subset of parameters are recorded for final reporting^[Model code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/inst/bi/BaseLineModel.bi].

@@

6. Manual independent proposal adaption at the approximate mean of the posterior distribution. It is assumed that the proposal for each parameter is guassian, truncated by the range of the prior with the mean based on the current parameter value. The standard deviation of the proposal distribution is halved each iteration, with at most 5 iterations of adaption. The minimum target acceptance rate specified was 10% and the maximum was 20%. 250 PMCMC samples were used each time to estimate the acceptance rate. Proposal adaption was only carried out for the main model scenario, with other scenarios using this proposal.

7. SMC-SMC model fitting with 1000 initial parameter particles. Particle rejuvernation was set to trigger when the effective sample size decreased below 25%, with 10 MCMC steps used each time.

>>>>

In the previous chapter I outlined a mechanistic model of Tuberculosis (TB) transmission. Whilst this model made use of the best available evidence there remains a large degree of uncertainty regarding its structure and parameterisation. The majority of this uncertainty relates to the amount of TB transmission occurring in England. An approach to deal with this uncertainty is to fit the model to available observed data. Model fitting involves optimising over the available parameter space to return parameter sets that fit the data in some quantitative way "better" than other parameter sets.

This chapter details an approach to challenging a infectious disease model to data using the state-space model formulation and Bayesian model fitting techniques. It first outlines the infectious disease model discussed in the previous chapter as a state-space model, as well as detailing the data used for fitting the model, and the parameters that are fitted. It then outlines the theoretical, and practical, justification for the model fitting pipeline used to calibrate and fit this state space model. Finally it discusses the quality of the model fit, ad-hoc techniques used to improve model fit, strengths and limitations of the approach, and areas for further work.

@@

State space models (SSMs) may be used to model dynamic systems that are partially observed (such as all but the most contained infectious disease outbreak or endemic). They consist of a set of parameters, a latent continuous/discrete time state process and an observed continuous/discrete time process.[@Murray2015] The model developed in the previous section represents the state process of the SSM, with the parameters estimated for the model representing the model initial conditions and parameter set. To complete the mapping to an SSM an observational model is required. This observational model takes the latent estimates from the dynamic model and forecasts the observed data. We specify such an observational model in the Section \@ref(observational-model).

@@

The primary data source used was the reported UK born TB notifications from 2000 to 2004 as recorded by the Enhanced TB Surveillance (ETS) system (see Chapter \@ref(data)). These are the only years for which notifications are available stratified by UK birth status and for which universal school-age BCG vaccination was in place. As the model of TB transmission outlined in the previous chapter had multiple age-dependent parameters UK born TB notifications stratified by age group are also fitted to. The age groups considered are children (aged 0-14 years old), adults (15-69 years old), and older adults (70-89 years old). These age groups were used, rather than the more detailed 5 year age groups used in the TB transmission model (see Chapter \@ref(model-development)), for three reasons. Firstly, by using condensed age groups the number of notifications in each group increased. Having a larger number of notifications in each group reduces the impact of stochastic noise, which makes fitting the model easier. Secondly, reducing the number of data points, whilst still capturing the important age dynamics, reduces the compute requirements of the model (see Section \@ref(particle-filter)). As will be discussed in Section \@ref(fitting-pipeline) this was a major consideration as the model fitting approached used was highly compute intensive. Finally, the condensed age groups used were chosen to be meaningful. This allows model results to be more easily reported and interpreted.

@@

There are three major considerations to account for when developing an observed disease notification model (i.e a reporting model). These are: systematic reporting error over time; systematic changes in reporting error over time; and reporting noise. For the notification data outlined in the previous section there is little evidence to suggest the form that this measurement model should take. For this reason I have assumed that all reporting errors are Gaussian and that their are no time variable reporting errors. The reporting model can be then defined as follows,

@@

Where $O$ are the observed notifications, $A$ are the incident cases of disease as forecast by the disease transmission model, $E_{\text{syst}}$ is the systematic reporting error, $E_{\text{noise}}$ is the reporting noise, and $\mathcal{N}$ represents the Gaussian (normal) distribution. The priors for the model are defined in Table \@ref(tab:measurement-model). The prior for systematic reporting error is based on the assumption that cases are less likely to be reported than they are to be over reported. The prior for the reporting noise is an assumption. This observation model was also used for the non-UK born pulmonary cases.

A potential limitation of this model is that reporting of TB cases is likely to have improved over time. This is especially true of notifications reported prior to the introduction of the ETS in 2000 (see Chapter \@ref(data)). A potential improvement to this model would be to introduce separate systematic reporting errors for notifications pre and post the introduction of the ETS. However, given that I have only included a single point prior to the introduction to the ETS doing so in this instance would lead to over-fitting of these parameters. Instead I have assumed that $ E_{\text{noise}}$ is doubled for data points prior to the introduction of the ETS.

@@

The model outlined in the Chapter \@ref(model-development) has a large number of free parameters for which prior distributions have been specified based on the observed data, the literature, and expert knowledge. In theory the model fitting pipeline outlined below could be used to produce posterior distributions for all these parameters. However, in practice this is not feasible as the data discussed in Section \@ref(observed-data) only covers notifications and therefore does not contain sufficient information. If every parameter was allowed to update based on the data then it is likely that the resulting posterior distributions would not match with alternative data sources and the literature. Another potential issue is that by allowing all parameters to be fitted the meaningful information in the observed data may be lost.

@@

In addition for scenarios with age variable transmission probabilities or non-UK born mixing the following parameters may also be fitted to,

* Transmission probability modifier for young adults ($\beta_{\text{young adult}}$).
* Non-UK born mixing modifier for young adults ($M_{\text{young adult}}$).

@@

Fitting dynamic transmission models is complex and requires the use of specialist statistical techniques. There are a variety of these tools available. Ranging for tried and tested to cutting edge. Historically many modellers have used maximum likelihood methods to fit deterministic models. Unfortunately, whilst computationally efficient, these methods only provide point estimates and are limited to relatively simple models. More recently Bayesian methods have become popular. These have numerous benefits including: explicit inclusion of prior knowledge via prior distributions for all parameters; ability to handle complex stochastic models; and provide parameter distributions (posterior distribution) of best fitting parameters rather than single point estimates. Unfortunately these methods also require calibration prior to use. This section outlines the theoretical justification, and implementation details, of an automated model fitting pipeline used to calibrate, and fit, the previously detailed state space model of TB transmission in England.

Libbi was used for all model fitting.[@Murray2015] LibBi is a software package for state-space modelling and Bayesian inference. It uses a domain specific language for model specification, which is then optimised and compiled to provide highly efficient model code. It focuses on full information model fitting approaches including: particle Markov chain Monte Carlo (PMCMC), and SMC-SMC methods for parameter estimation. All fitting algorithms are highly efficient and scalable across multiple CPUs or GPUs. `RBI` and `RBI.helpers` were used to interface with LibBi from R.[@Funk:2019ud; @Funk:2019uw] `RBI.helpers` was also used to optimise the model fitting pipeline as detailed in the calibration section. As model fitting using `LibBi` is compute intensive a workstation was built, and overclocked, with these compute requirements in mind^[See these blog posts for details: https://www.samabbott.co.uk/post/building-an-rstats-workstation/, https://www.samabbott.co.uk/post/benchmarking-workstation-xgboost/, https://www.samabbott.co.uk/post/benchmarking-workstation-benchmarkme/]. All model fitting code is available on GitHub as an R package^[See here for details: https://github.com/seabbs/ModelTBBCGEngland].

@@

In order to fit a model to data it is necessary to estimate, or calculate, the marginal likelihood. Mathematically, the marginal likelihood is the plausibility that a parameter set, given the specified statistical model and the initial conditions, describes the observed data. For complex state space models, such as that discussed in the previous chapter, calculating the marginal likelihood is not possible.[@Murray2015] The particle filter provides a model-agnostic approach, based on importance sampling, to estimate the marginal likelihood. The variant used in this thesis, the bootstrap particle filter, is described below. See [@Murray2015] for a more technical discussion of the bootstrap particle filter.

@@

2. Sequentially for the each observed data point the particle filter is then advanced through a series of *propagation*, *weighting*, and *re-sampling* steps.

@@

The particle filter approach outlined above, is a member of a family of sequential Monte Carlo (SMC) methods. These methods all initialise particles and then follow the same *propagation*, *weighting*, and *re-sampling* steps as previously detailed. SMC may also be used to sample from the posterior distribution of a given set of priors and a specified model. This works as follows,

@@

2. These parameter particles are then iterated sequentially over each observed data point, undergoing the same *propagation*, *weighting*, and *re-sampling* steps as in the particle filter, as well as an additional *rejuvenation* step.

@@

   * *Weighting:* Parameter particles are weighted using the marginal likelihood. In principle this could be computed exactly, but is most commonly estimated using a nested particle filter for each state particle (i.e as outlined in the previous section). For a subset of models a Kalman filter maybe used instead.[@Murray2015] The marginal likelihood may also be estimated using other partial information techniques such as approximate Bayesian computation. In the case where a particle filter is used the full algorithm is known as Sequential Monte Carlo - Sequential Monte Carlo (SMC-SMC). This algorithm is used for all dynamic model fitting in this thesis.

 *  *Re-sampling:* The parameter particle stock is restored to equal weights by re-sampling particles, with replacement, with the probability of each sample being drawn being proportional to its weight.

 * *Rejuvenation:* *Re-sampling* of the parameter particles at each time point leads to a reduction in the number of unique values present. For state particles (when estimating the marginal likelihood using a particle filter) particles are diversified with each propagation but as parameters do not change in time parameter particles cannot diversify in this way. To account for this the *rejuvenation* step is inserted after the re-sampling of parameter particles at each time point. The *rejuvenation* step is a single, or multiple depending on the acceptance rate, Metropolis-Hastings step for each parameter particle. This step aims to preserve the distribution of the parameter particles, whilst increasing their diversity. To minimize unnecessary rejuvenation an effective sample size threshold can be used. This only triggers rejuvenation when particle diversity has decreased below the target effective sample size threshold.

@@

Where $y$ is the observed data, $\theta$ is the current parameter set, $\theta'$ is some proposed parameter set sampled from some proposal distribution $q(\theta' | \theta)$.[@Murray2015] By construction samples drawn using this rule are ergodic to the posterior distribution. This means that after convergence samples drawn using this rule may be considered as samples from the posterior distribution.

@@

The accuracy of the marginal likelihood estimate returned by the particle filter is dependent on the number of particles used, the number of observed data points, the parameter sample, and the complexity of the model. As the number of particles tends towards infinity the likelihood estimate provided by the particle filter should tend towards the exact solution. This suggests that choosing a very high number of particles maybe the optimal solution in terms of accuracy. Unfortunately, each particle requires a full model simulation, which for complex models can be computationally costly. This means that using very large numbers of particles is not tractable. For this reason it is necessary to determine an optimal number of particles that both provides an adequately accurate estimate of the likelihood whilst being computationally tractable.

The `rbi.helpers` R package attempts to solve this issue by adopting the following strategy.[@Funk:2019uw] First, the approximate mean of the posterior distribution is obtained, as accurate likelihood estimates near the posterior mean are of the most interest. Repeated model simulations are then run using the same set of parameters, with the marginal likelihood being estimated each time using a given number of particles. The variance of these log-likelihood estimates is then calculated. This process is then repeated for increasing numbers of particles until the log-likelihood variance is below some target threshold, commonly 1.[@Funk:2019uw]

I have implemented this as a two step process for each fitted scenario. Firstly, I used the Nelder-Mead simplex method, via LibBi,[@Murray2015] to find a parameter set that optimised the maximum likelihood. I then initialised a 1000 step PMCMC chain with this parameter set, using 256 particles in the particle filter. I then used `rbi.helpers`,[@Funk:2019uw] as outlined above, to estimate the number of particles required to produce a log-likelihood variance of less than 1 for this sample of the posterior distribution, using 250 samples per step and starting with 16 particles. I initially planned to repeat this process for multiple draws from the posterior distribution but this proved to be in-feasible given the compute available. A target of 5 for the log-likelihood variance was chosen as a smaller target could not be feasibly achieved given the compute resources available. Additionally 256 was specified as the maximum number of feasible particles to use in the particle filter.

@@

When using an MCMC algorithm a proposal distribution is required to provide new parameter samples to evaluate. For SMC-SMC a proposal distribution is required to inform the MCMC sampler that is run during the rejuvenation step. By default if no proposal distribution is provided Libbi uses the prior distribution.[@Murray2015] The prior distribution can be an inefficient proposal distribution as it is likely to have a low acceptance rate (from the MCMC sampler).[@Murray2015] Having a low acceptance rates means that many more MCMC steps are required to generate a successful parameter sample. This results in slow mixing and computationally expensive MCMC steps may make model fitting intractable.

A more efficient approach is to specify a proposal distribution that draws parameter samples that are closer to the current state of the MCMC chain than the overall prior distribution. There is an extensive literature examining how to optimise the proposal distribution to achieve an good acceptance rate. In practice it has been shown that a rate of between 10% and 20% is optimal for upwards of 5 parameters.[@Murray2015] This strikes a balance between allowing the chain to fully explore the posterior distribution whilst still being as efficient as possible.

A simple approach to setting the proposal is to run a series of MCMC steps and then calculate the acceptance rate. Based on the acceptance rate the width of the proposal distributions can then be adapted. By repeating these steps multiple times a proposal distribution which gives an acceptance rate within the desired bounds can be arrived at. This adaption can either be independent for each parameter or dependent (taking into account empirical correlations). The `adapt_proposal` function, from the `rbi.helpers` R package,[@Funk:2019uw] implements this approach and is used in this model fitting pipeline. In many models parameters are likely to have strong correlations (i.e between UK and Non-UK born mixing rate and effective contact rate) and in these scenarios it is likely that a dependent strategy for adapting the proposal distribution will more efficiently explore the posterior distribution. However, the downside of adapting the proposal distribution using dependent methods is that the resulting proposal is highly complex, is computationally expensive to compute and may breakdown in some areas of the posterior distribution.

In this model fitting pipeline I have used a maximum of 5 iterations of, manual, independent proposal adaption, drawing 250 samples in each iteration, starting with Gaussian distributions for each parameter, truncated by the range of the prior, with the mean based on the current parameter value. The standard deviation for each parameter was assumed to be the standard deviation of the prior if it was Gaussian and otherwise assumed to be the range of the prior if it was uniform. For each iteration I halved the  size of the standard deviation of each parameter. As for the particle calibration, I initially used a maximum likelihood method to provide a point estimate of the best fitting parameter set, followed by 1000 PMCMC steps, using a 256 particle filter. This means that the proposal distribution is adapted near to the posterior mean rather than in the tails of the posterior distribution.

I chose to use, manual, independent proposal adaption method for several reasons. Firstly, when developing this pipeline the approaches implemented in `rbi.helpers` produced multiple transient errors in other `rbi` and `rbi.helpers` code. Secondly, the resulting dependent proposal distribution was highly complex, slow to compute, and difficult to debug. Finally, for SMC-SMC efficient exploration of the proposal distribution is less important than when using MCMC alone as SMC-SMC is initialised with multiple samples from the prior distribution. This means that multiple local maximas can be efficiently explored regardless of the proposal distribution used. The MCMC rejuvenation step then serves to provide additional samples from these local maximas. Proposal adaption was only carried out for the main model scenario with all other scenarios using this proposal distribution.

### Model comparison

@@

Smaller DIC values should indicate a better fit to data than larger DIC values. The DIC is composed of the deviance, which favours a good fit, and the effective number of parameters, which penalises over-fitting.[@Gelman:nll_LBlw] Unlike the AIC the DIC can be estimated using samples from the posterior distribution and so is more readily calculated for models estimated using Bayesian methods. It can be defined as,

$$ {\mathit  {DIC}}=D({\bar  {\theta }})+2p_{D}$$
Where $\bar{\theta}$ is the expectation of $\theta$, with $\theta$ being defined as the unknown parameters of the model. $p_{D}$ is the effective number of parameters in the model and is used to penalise more  complex models. It can be estimated as follows,[@Gelman:nll_LBlw]

@@

PRCC is a sampling based approach which can be computed with minimal computational cost from a sample of the prior or posterior distributions of a model. It estimates the degree of correlation between a given parameter input and an output after adjusting (using a linear model) for variation in other inputs. It is an extension of more simplistic sampling techniques, the most basic of which, is simply examining scatter plots of a sampled parameter set against the outcome of interest. PRCC is required as these more simplistic techniques become intractable with higher dimensionality as they do not account for between parameter correlation or are just difficult to interpret with multiple dimensions.[@Marino2009a] PRCC can be understood by first outlining the individual steps. These are:

@@

Where $\operatorname {cov}$  is the co-variance, $\sigma _{X}$ is the standard deviation of $X$, and $\sigma_Y$  is the standard deviation of $Y$. Where $X$ is the input and $Y$ is the output.

@@

A limitation of PRCC is that it whilst it can capture non-linear relationships between outputs and inputs these relationships must be monotonic.[@Marino2009a] For relationships that are non-monotic methods that rely on the decomposition of model output variance, such as the extended Fourier amplitude sensitivity test,[@Marino2009a] are more appropriate. However, these approaches are computationally demanding as they typically require multiple iterations of model simulation. Additionally, they cannot be used on a previous parameter samples, instead needing to sample and simulate the model within the parameter sensitivity algorithm. This means that they cannot be used for "free" (i.e with negligible additional compute cost), unlike PRCC which can be estimated using a sample from the posterior distribution. For this reason these approaches have not been further explored in this thesis.

I have implemented PRCC using the `epiR` R package^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/test_sensitivity.R],[@EpiR] using the samples from the posterior distribution of the model calculated during the SMC-SMC step. Parameter sensitivity measures such as PRCC must be calculated separately for each output time point. I calculated the PRCC for each fitted parameter, at the final time point with fitted data (2004), for overall TB incidence rates. These results are then summarised by plotting the absolute PRCC values, indicating the direction of correlation using colour^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/plot_sensitivity.R].

@@

1. Model initialisation using minimal error checking and single precision computation. Implemented using the `disable-assert` and `enable-single` flags in LibBi.[@Murray2015] Outputs are only given for times with observed data and a subset of parameters are recorded for final reporting^[Model code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/inst/bi/BaseLineModel.bi].

@@

6. Manual independent proposal adaption at the approximate mean of the posterior distribution. It is assumed that the proposal for each parameter is Gaussian, truncated by the range of the prior with the mean based on the current parameter value. The standard deviation of the proposal distribution is halved each iteration, with at most 5 iterations of adaption. The minimum target acceptance rate specified was 10% and the maximum was 20%. 250 PMCMC samples were used each time to estimate the acceptance rate. Proposal adaption was only carried out for the main model scenario, with other scenarios using this proposal.

7. SMC-SMC model fitting with 1000 initial parameter particles. Particle rejuvenation was set to trigger when the effective sample size decreased below 25%, with 10 MCMC steps used each time.
